---
title: "Gov 51 - Pset 1"
author: "Tyler Dang - https://github.com/tylerdang25/Gov-51---Pset-1.git"
format: pdf
editor: visual
---

## 2. Get to Know Your Data

```{r}
library(tidyverse)
library(dplyr)
library(knitr)
```

```{r}
acs2024 <- read.csv('/Users/tyler/Documents/Gov 51/gov51-ps1/data/raw/acs2024.csv')
```

FROM THE DOCUMENTATION:

TRANTIME reports the total amount of time, in minutes, that it usually took the respondent to get from home to work last week.

A 000 for TRANTIME represents cases in which there is no data available for that observation.

PERWT indicates how many persons in the U.S. population are represented by a given person in an IPUMS sample. Basically, for each observation in this data set, the PERWT value indicates how many real people in the U.S. population are represented by that observation.

```{r}
acs2024 <- acs2024 %>% 
  mutate(female = ifelse(SEX == 2, 1, 0))

acs2024 <- acs2024 %>% 
  mutate(educ_lhs = ifelse(EDUC %in% 1:5, 1, 0),
    educ_hs = ifelse(EDUC == 6, 1, 0),
    educ_somecol = ifelse(EDUC %in% 07:09, 1, 0),
    educ_college = ifelse(EDUC == 10, 1, 0),
    educ_adv = ifelse(EDUC == 11, 1, 0))

acs2024 <- acs2024 %>% 
  mutate(employed = ifelse(EMPSTAT == 1, 1, 0),
    unemployed = ifelse(EMPSTAT == 2, 1, 0),
    nlf = ifelse(EMPSTAT == 3, 1, 0))
```

If we included special codes that represent "N/A" codes, we may accidentally misreport in summary statistics the extent that a certain population appears. For example, if we included 0 as a part of the range for education less than high school, we would drastically increase the extent that those with an education level less than high school make up the data set instead of accurately showing that the information may just be missing.

```{r}

acs2024table <- data.frame(
  Variable = c("Age", "Female", "Less than High School", "High School Only", 
               "Some College", "College Only", "Advanced Degree", "Employed", 
               "Unemployed", "Not in Labor Force", "Commute Time (mins)", 
               "Total Income ($)"),
  
  N = c(length(acs2024$AGE), length(acs2024$female),length(acs2024$educ_lhs), 
        length(acs2024$educ_hs), length(acs2024$educ_somecol), 
        length(acs2024$educ_college), length(acs2024$educ_adv),
        length(acs2024$employed), length(acs2024$unemployed),
        length(acs2024$nlf), length(acs2024$TRANTIME), length(acs2024$INCTOT)),
  
  Mean = c(mean(acs2024$AGE), mean(acs2024$female), mean(acs2024$educ_lhs), 
           mean(acs2024$educ_hs), mean(acs2024$educ_somecol), 
           mean(acs2024$educ_college), mean(acs2024$educ_adv),
           mean(acs2024$employed), mean(acs2024$unemployed),mean(acs2024$nlf),
           mean(acs2024$TRANTIME), mean(acs2024$INCTOT) / 1e6),
  
  SD = c(sd(acs2024$AGE), sd(acs2024$female),sd(acs2024$educ_lhs), 
         sd(acs2024$educ_hs), sd(acs2024$educ_somecol), 
         sd(acs2024$educ_college), sd(acs2024$educ_adv), sd(acs2024$employed), 
         sd(acs2024$unemployed),sd(acs2024$nlf), sd(acs2024$TRANTIME), 
         sd(acs2024$INCTOT) / 1e6),
  
  Min = c(min(acs2024$AGE), min(acs2024$female), min(acs2024$educ_lhs), 
          min(acs2024$educ_hs), min(acs2024$educ_somecol), 
          min(acs2024$educ_college), min(acs2024$educ_adv),
          min(acs2024$employed), min(acs2024$unemployed), min(acs2024$nlf),
          min(acs2024$TRANTIME), min(acs2024$INCTOT) / 1e6),
  
  Max = c(max(acs2024$AGE), max(acs2024$female),max(acs2024$educ_lhs), 
          max(acs2024$educ_hs), max(acs2024$educ_somecol), 
          max(acs2024$educ_college), max(acs2024$educ_adv),
          max(acs2024$employed), max(acs2024$unemployed), max(acs2024$nlf),
          max(acs2024$TRANTIME), max(acs2024$INCTOT) / 1e6)
)

#acs2024table
acs2024rounded <- acs2024table
acs2024rounded[, -1] <- round(acs2024table[, -1], 2)
#acs2024rounded

knitr::kable(acs2024rounded, 
             caption = "Table 1: Summary Statistics for 2024 ACS Sample",
             digits = 2)
```

The maximum daily commute time of 195 minutes or 3.25 hours can make sense for a daily commute. The statistic means that the max reported commute from home to work was 195 minutes, and though this is not very practical for most workers, it is possible for the commute to be made daily. However, with a commute of more than 3 hours, that is 6 total hours daily in commute which leaves little time for rest and other necessary actions in the course of one's day.

## 3. Who Should Be in Your Analysis

```{r}
```
